---
title: "Balanced Incomplete Block Experiment"
---

# Balanced incomplete block experiment

## Background

The block design in **link RCBD guide** was complete, meaning that every block contained all the treatments. In practice, it may not be possible to have too many treatments in each block. Sometimes, there are also situations where it is advised to not have many treatments in each block.

In such cases, incomplete block designs are used where we have to decide what subset of treatments to be used in an individual block. This will work well if we enough blocks. However, if we only have small number of blocks, there would be the risk that certain quantities are not estimable anymore.

To avoid having a disconnected design, a balanced incomplete block design can be used 

::: callout-note
## A note


:::


## Example Analysis

https://kwstat.github.io/agridat/reference/weiss.incblock.html

```{r}
 library(agridat)
  data(weiss.incblock)
  dat <- weiss.incblock
```

```{r}
 library(desplot)
  desplot(dat, yield~col*row,
          text=gen, shorten='none', cex=.6, out1=block,
          aspect=252/96, # true aspect
          main="weiss.incblock")
  
```

::: {.panel-tabset}

### lme4

```{r, message=FALSE}

```

### tidymodels

```{r, message=FALSE}

```


:::




|   |   |
|----------|----------------------------------------|    


### *Data integrity checks*



### Model Building


::: {.column-margin}

Recall the model:

$$ $$ 

:::

Here is the R syntax for that statistical model:

::: {.panel-tabset}

### lme4

```{r}

```

### tidymodels

```{r}

```

:::



### Check Model Assumptions


### Inference
