---
title: "Incomplete Block Design"
---

```{r, include=FALSE, echo=FALSE}
source(here::here("settings.r"))

par(mar=c(5.1, 6, 4.1, 2.1))
```

## Background

The block design in [**Chapter 4**](rcbd.qmd) was complete, meaning that every block contained all the treatments. In practice, it may not be possible to have too many treatments in each block. Sometimes, there are also situations where it is advised to not have many treatments in each block.

In such cases, incomplete block designs are used where we have to decide what subset of treatments to be used in an individual block. This will work well if we enough blocks. However, if we only have small number of blocks, there would be the risk that certain quantities are not estimable anymore.

Incomplete block designs are grouped into balanced lattice design and partially balanced (or alpha-lattice) designs. 

To avoid having a disconnected design, a balanced incomplete block design can be used

The statistical model for balanced incomplete block design is:

$$y_{ij} = \mu + \alpha_i + \beta_j + \epsilon_{ij}$$ Where:

$\mu$ = overall experimental mean $\alpha$ = treatment effects (fixed) $\beta$ = block effects (random) $\epsilon$ = error terms

$$ \epsilon \sim N(0, \sigma)$$

$$ \beta \sim N(0, \sigma_b)$$ There are few key points that we need to keep in mind while designing incomplete block designs:

-   A drawback of this design is that block effect and treatment effects are confounded.

-   To eliminate the block effects, better compare treatments within a block.

-   No treatment should appear twice in any block as they contributes nothing to within block comparisons.

An [excellent description of incomplete block design](https://people.math.ethz.ch/~meier/teaching/anova/incomplete-block-designs.html) is provided in [*ANOVA and Mixed Models*](https://www.taylorfrancis.com/books/mono/10.1201/9781003146216/anova-mixed-models-lukas-meier) by Lukas Meier. 


The balanced incomplete block designs are guided by strict principles and guidelines including: the number of treatments must be a perfect square (e.g. 25, 36, and so on); number of replicates must be equal to no. of blocks +1;  

::: callout-note
## A note

Because the blocks are incomplete, the Type I and Type III sums of squares will be different. That is, the missing treatments in each block represent missing observations (but not missing ‘at random’).
:::

## Balanced Incomplete Block Design
### Example Analysis
We will demonstrate an example data set designed in a balanced incomplete block design. First, load the libraries required for analysis and estimation.

::: panel-tabset
### lme4

```{r, message=FALSE, warning=FALSE}
library(lme4); library(lmerTest); library(emmeans)
library(dplyr); library(broom.mixed); library(performance)
```
### nlme
```{r, message=FALSE, warning=FALSE}
library(nlme); library(broom.mixed); library(emmeans)
library(dplyr); library(performance)
```
:::

The data used for this example analysis was extracted from the `agridat` package. This example is comprised of soybean balanced incomplete block experiment.


```{r}
dat <- agridat::weiss.incblock
```
|       |                              |
|-------|------------------------------|
| block   | blocking unit             |
| gen   | genotype (variety) factor                |
| row   | row position for each plot        |
| col   | column position for each plot    |
| yield | grain yield in bu/ac         |

: Table of variables in the data set {tbl-incom_blk}

```{r, fig.height= 9, echo=FALSE}

desplot::desplot(dat, yield~col*row,
          text=gen, shorten='none', cex=.6, out1=block,
          aspect=252/96, # true aspect
          main="weiss.incblock")
```


#### Data integrity checks

We will start inspecting the data set firstly by looking at the class of variables:

```{r}
str(dat)
```

The variables we need for the model are block, gen  and yield. The block and gen are classified as factor variables and yield is numeric. Therefore, we don't need to change class of any of the required variables.

Next, let's check the independent variables. We can look at this by running a cross tabulations among block and gen factors.

```{r}
table(dat$block, dat$gen)
```

```{r}
agg_tbl <- dat %>% group_by(gen) %>% 
  summarise(total_count=n(),
            .groups = 'drop')
agg_tbl
```

```{r}
agg_df <- aggregate(dat$gen, by=list(dat$block), FUN=length)
agg_df
```


There are 31 varieties (gen) and it is perfectly balanced, with exactly one observation per treatment per block.

We can calculate the sum of missing values in variables in this data set to evaluate the extent of missing values in different variables

```{r}
apply(dat, 2, function(x) sum(is.na(x)))
```
We observed no missing data!

Last, let's plot a histogram of the dependent variable. This is a quick check before analysis to see if there is any strong deviation in values.
```{r, echo=FALSE}
#| label: fig-ibd_hist
#| fig-cap: "Histogram of the dependent variable."
#| column: margin
par(mar=c(5.1, 5, 2.1, 2.1))
hist(dat$yield, main = "", xlab = "yield", cex.lab = 1.8, cex.axis = 1.5)
```

```{r, eval=FALSE}
hist(dat$yield, main = "", xlab = "yield")
```
Response variable values fall within expected range, with few extreme values on right tail.
This data set is ready for analysis!

#### Model Building
We will be evaluating the response of yield as affected by gen (fixed effect) and block (random effect).

::: column-margin
Please note that incomplete block effect can be analyzed as a fixed (intra-block analysis) or a random (inter-block analysis) effect. When we consider block as a random effect, the mean values of a block also contain information about the treatment effects.
:::
::: panel-tabset
### lme4

```{r}
model_icbd <- lmer(log(yield) ~ gen + (1|block),
                   data = dat, 
                   na.action = na.exclude)
tidy(model_icbd)
```

### nlme
```{r}
model_icbd1 <- lme(yield ~ gen,
                  random = ~ 1|block,
                  data = dat, 
                  na.action = na.exclude)
tidy(model_icbd1)
```
:::

#### Check Model Assumptions
Let's verify the assumption of linear mixed models including normal distribution and constant variance of residuals. 

::: panel-tabset
#### lme4
```{r, fig.height=3}
check_model(model_icbd, check = c('normality', 'linearity'))
```

#### nlme
```{r, fig.height=3}
check_model(model_icbd1, check = c('normality', 'linearity'))
```
:::
::: column-margin
Here we observed a right skewness in residuals, this can be resolved by using data transformation e.g. log transformation of response variable. Please refer to **chapter** to read more about data transformation. 
:::

#### Inference

We can extract information about ANOVA using `anova()` from lmer and lme models, respectively. 
::: panel-tabset
#### lme4
```{r}
anova(model_icbd, type = "1")
```

#### nlme
```{r}
anova(model_icbd1, type = "sequential")
```
:::

Let's look at the estimated marginal means of yield for each variety (gen).
::: panel-tabset
#### lme4
```{r}
emmeans(model_icbd, ~ gen)
```

#### nlme
```{r}
emmeans(model_icbd1, ~ gen)
```
:::


##data2

The data used in this example is extracted from the `agridat` package. This data is a balanced lattice experiment in cotton containing 16 treatments in a 4x4 layout in each of 5 replicates. The response variable in this data is the percentage of young flower buds attacked by boll weevils.
```{r}
dat3 <- agridat::cochran.lattice
#head(dat3)
```
|       |                              |
|-------|------------------------------|
| rep   | replication unit             |
| trt   | treatment factor                |
| row   | row position for each plot        |
| col   | column position for each plot    |
| y | % of young flower buds attacked        |

: Table of variables in the data set {tbl-lattice}


```{r}
str(dat3)
table(dat3$trt, dat3$rep)
```

```{r}
library(desplot)
desplot(dat3, y~row*col|rep,
        text=trt, # aspect unknown, should be 2 or .5
         main="cochran.lattice")

```


Here, we can use the `desplot()` function from the 'desplot' package to visualize the plot plan from lattice design.

```{r}
desplot::desplot(dat3, y~row*col|rep,
                 text=trt, shorten='none', cex=.6,
                 aspect=252/96, # true aspect
                 main="Balanced incomplete block")

```

### Data integrity checks
```{r, echo=FALSE}
#| label: lattice_design
#| fig-cap: "Histogram of the dependent variable."
#| column: margin
par(mar=c(5.1, 5, 2.1, 2.1))
desplot(dat3,  y~row*col|rep,
                 text=trt, shorten='none', cex=.6,
                 aspect=252/96, # true aspect
                 main="Balanced incomplete block")
```
### Data integrity checks

```{r}
str(dat3)
```

```{r}
#dat2$row <- as.factor(dat2$row)
#dat2$col <- as.factor(dat2$col)

dat3$row <- as.factor(dat3$row)
dat3$col <- as.factor(dat3$col)
```

```{r, eval=FALSE}
hist(dat3$y, main = "", xlab = "yield")
```

```{r, echo=FALSE}
#| label: fig-rcbd_hist
#| fig-cap: "Histogram of the dependent variable."
#| column: margin
par(mar=c(5.1, 5, 2.1, 2.1))
hist(dat3$y, main = "", xlab = "yield", cex.lab = 1.8, cex.axis = 1.5)
```

::: panel-tabset
### lme4

```{r}
m1_a <- lmer(y ~ trt + (1|rep) + (1|rep:row:rep) + (1|rep:col),
           data = dat3,
           na.action = na.exclude)
tidy(m1_a) 
```

### nlme

```{r}
dat3$dummy <- factor(1)

m1_b <- lme(y ~ trt,
          random = list(dummy = pdBlocked(list(
                                  pdIdent(~rep - 1),
                                  pdIdent(~rep:row - 1),
                                  pdIdent(~rep:col)))),
          data = dat3, 
          na.action = na.exclude)

VarCorr(m1_b)
```
:::

### Check Model Assumptions

Remember those iid assumptions? Let's make sure we actually met them.

```{r, fig.height=5}
check_model(m1_a, check = c("linearity", "normality"))
```


### Inference

Estimates for each treatment level can be obtained with the 'emmeans' package. And we can extract the ANOVA table from model using `anova()` function.

```{r}
anova(m1_a)
```

Estimated marginal means

```{r}
emmeans(m1_a, ~ trt)
```

## Aplha Lattice Design (partially-balanced)
### Example Analysis

In incomplete block alpha-design, the blocks are grouped into complete replicates. These designs are also termed as "resolvable incomplete block designs" or "partially balanced incomplete block designs"[^analysis-tips-1]. This design has been more commonly used instead of balanced IBD because of it's practicability, flexibility, and versatility. 
[^analysis-tips-1]:Patterson, H. D., & Williams, E. (1976). A new class of resolvable incomplete block designs. Biometrika, 63(1), 83-92.

The data used in this example is published in John and Williams (1995)[^analysis-tips-2]. The data in this trial was laid out in an alpha lattice design. This trial data had 24 genotypes (gen), 6 incomplete blocks, and each replicated 3 times. 

[^analysis-tips-2]: Hering, F. (1996). John, JA, Williams, ER: Cyclic and Computer Generated Designs. Chapmann and Hall, London, New York 1995, 255 S., $ L 32,‐.


Let's start analyzing this example firstly by loading the required libraries for linear mixed models:

::: panel-tabset
### lme4
```{r, message=FALSE, warning=FALSE}
library(lme4); library(lmerTest); library(emmeans)
library(dplyr); library(broom.mixed); library(performance)
```
### nlme
```{r, message=FALSE, warning=FALSE}
library(nlme); library(broom.mixed); library(emmeans)
library(dplyr); library(performance)
```
:::

#### Data integrity checks

```{r}
 base::library(readr)
dataURL <- "https://raw.githubusercontent.com/SchmidtPaul/DSFAIR/master/data/John%26Williams1995.csv"
dat <- read_csv(dataURL)

dat
```

```{r}
dat <- dat %>% 
  mutate_at(vars(plot:gen), as.factor)

```

```{r}
desplot::desplot(data = dat, flip = TRUE,
        form = gen ~ col + row | rep,          # fill color per genotype, headers per replicate
        text = gen, cex = 0.7, shorten = "no", # show genotype names per plot
        out1 = rep,                            # lines between complete blocks/replicates
        out2 = inc.block,                      # lines between incomplete blocks
        main = "Field layout", show.key = F)   # formatting
```


https://kwstat.github.io/agridat/reference/john.alpha.html 
```{r}
dat4 <- agridat::john.alpha
```
|       |                              |
|-------|------------------------------|
| block   | blocking unit             |
| gen   | genotype (variety) factor                |
| row   | row position for each plot        |
| col   | column position for each plot    |
| yield | grain yield in bu/ac         |

: Table of variables in the data set {tbl-incom_blk1}





