---
title: "lattice_design"
---

# Background

Yates (1936) proposed this method of arranging agricultural variety trials involving a large number of crop varieties. These types of arrangements were named a quasi-factorial or lattice designs. His paper contained numerical examples based on the results of a uniformity trial on orange trees. A special feature of lattice designs is that the number of treatments, t, is related to the block size, k, in one of three forms: t = k\^2, t = k3, or t = k(k +1).

Even though the number of possible treatments is limited, a lattice design may be an ideal design for field experiments with a large number of treatments.

https://kwstat.github.io/agridat/reference/cochran.lattice.html


Statistical model for lattice design:

$Yijk = \mu + \alpha_i + \gamma_j + \tau_t  + \beta_k + \epsilon_ijk$

where, $\mu$ is the experiment mean, 𝛽 is the row effect, 𝛾 is the column effect, and 𝜏 is the treatment effect.

## Example Analysis

First, load the libraries for analysis and estimation:

::: panel-tabset
### lme4
```{r, message=FALSE, warning=FALSE}
library(lme4); library(lmerTest); library(emmeans); library(performance)
library(dplyr); library(broom.mixed); library(agridat); library(desplot)
```
### nlme
```{r, message=FALSE, warning=FALSE, eval=FALSE}
library(nlme); library(broom.mixed); library(emmeans); library(performance)
library(dplyr); library(agridat); library(desplot)
```
:::

Import data from agridat package and create columns for row and column as factor variables. This is a balanced experiment design

```{r}
data(burgueno.rowcol)
dat <- burgueno.rowcol
```

```{r}
# Two contiguous reps in 8 rows, 16 columns

desplot(dat, yield ~ col*row,
        out1=rep, # aspect unknown
        text=gen, shorten="none", cex=.75,
        main="burgueno.rowcol")

```
### Data integrity checks
```{r}
str(dat)
```
```{r}
dat$row <- as.factor(dat$row)
dat$col <- as.factor(dat$col)
```

```{r, eval=FALSE}
hist(dat$yield, main = "", xlab = "yield")
```

```{r, echo=FALSE}
#| label: fig-rcbd_hist
#| fig-cap: "Histogram of the dependent variable."
#| column: margin
par(mar=c(5.1, 5, 2.1, 2.1))
hist(dat$yield, main = "", xlab = "yield", cex.lab = 1.8, cex.axis = 1.5)
```

::: panel-tabset
### lme4

```{r}
m1 <- lmer(yield ~ gen + (1|rep) + (1|rep:row) + (1|rep:col), 
           data=dat, 
           na.action = na.exclude)
summary(m1) 
```

### nlme

```{r, eval=FALSE}
## lme not working for this, need help in fixing it
#m1 <- lme(yield ~ gen,
          random = ~ 1|rep + 1|rep:row + 1|rep:col,
          data = dat, 
          na.action = na.exclude)
```
:::

### Check Model Assumptions

Remember those iid assumptions? Let's make sure we actually met them.
```{r, fig.height=9}
check_model(m1)
```



```{r}
#Random rep, row and col within rep
 m1 <- lmer(yield ~ gen + (1|rep) + (1|rep:row) + (1|rep:col), data=dat)
summary(m1) 
anova(m1)

```
### Inference

Estimates for each treatment level can be obtained with the 'emmeans' package. And we can extract the ANOVA table from model using `anova()` function.

```{r}
anova(m1)
```

Estimated marginal means

```{r}
emmeans(m1, ~ gen)
```
