---
title: Marginal Means & Contrasts
---

## Background

Estimated marginal means are the means of a variable averaged across all levels of other variables in a model, conditioned on the variance-covariance matrix. That is to say, these are much more than plain averages. The ***emmeans*** package is one of the most commonly used package in R for determining marginal means (also known as **least-squares means**). In this chapter, we will demonstrate the extended use of the ***emmeans*** package to calculate estimated marginal means and confidence intervals and conduct posthoc comparisons.

To demonstrate the use of the ***emmeans*** package. We will use the model from split plot lesson ([**Chapter 7**](split-plot-design.qmd)), where we evaluated the effect of nitrogen and variety on oat yield. This data contains six blocks, three main plots (Variety) and four subplots (Nitrogen). The dependent variable was oat yield. To read more about the experiment layout details please read RCBD split-plot section in [**Chapter 7**](split-plot-design.qmd).


::: callout-note
#### Marginal means using lmer and nlme

For demonstration of the `emmeans` package, we are fitting model with `nlme` package. Please note that code below calculating marginal means works for both `lmer()` and `nlme()` models.
:::

## Analysis Examples
Let's start this example by loading the required libraries for fitting linear mixed models using **nlme** package.

```{r, warning=FALSE, message=FALSE}
library(nlme); library(performance); library(emmeans)
library(dplyr); library(broom.mixed); library(multcompView)
library(multcomp); library(ggplot2)
```

### Import data

Let's import oats data from the MASS package.

```{r}
data1 <- MASS::oats
```

::: column-margin
To read more about data and model fitting explanation please refer to [**Chapter 7**](split-plot-design.qmd).
:::

### Model fitting
```{r}
model1 <- lme(Y ~  V + N + V:N ,
                  random = ~1|B/V,
                  data = data1, 
                  na.action = na.exclude)
tidy(model1)
```

### Check Model Assumptions
```{r,  fig.height=3}
check_model(model1, check = c('qq', 'linearity'), detrend=FALSE, alpha = 0)
```

Residuals look good with a small hump in middle and normality curve looks reasonable.

### Model Inference

```{r}
anova(model1, type = "marginal")
```
The analysis of variance showed a significant N effect and no effect of V and V x N interaction effect on oat yield.

### Estimated Marginal Means

Now that we have fitted a linear mixed model (model1) and it meets the model assumptions, we can use the `emmeans()` function to obtain estimated marginal means for main effects (variety and nitrogen) and interaction (variety x nitrogen). 

#### Main effects

The main effects in model1 were V and N. 
```{r, warning=FALSE}
m1 <- emmeans(model1, ~V, level = 0.95)
m1
```

```{r, warning=FALSE}
m2 <- emmeans(model1, ~N)
m2
```
Make sure to read and interpret marginal means carefully. Here, when we calculated marginal means for main effects of V and N, these were averaged over the levels of other factor in experiment. For example, estimated means for each variety were averaged over the N treatments.

#### Interaction effects

Now let's evaluate the marginal means for the interaction effect of V and N. These can be calculated either using `V*N` or `V|N`. 
```{r}
m3 <- emmeans(model1, ~V|N)
m3
```

```{r}
m4 <- emmeans(model1, ~V*N)
m4
```

:::{.callout-note}
Here, note that we estimated marginal means in two ways: 
`V*N` estimates the marginal means for each combination of levels across these varaibles, and when conducting pairwise comparisons on these, all combinations are compared to one another simultaneously (not recommended). 

`V|N` estimates the marginal means for each V at a given level of N. When conducting pairwise comparisons on these, all levels of the variable V are compared with each other within each level of N, not across all levels of N (a sensible approach). 
:::

## Contrasts using emmeans

The `pairs()` function from the emmeans package can be used to evaluate pairwise comparisons among treatment factors. The emmean object (m1, m2) will be passed through the `pairs()` function, which will provide a p-value adjustment equivalent to the Tukey test.

```{r}
pairs(m1, adjust = "tukey")
```

```{r}
pairs(m2)
```

The more levels in a variable, the more comparisons that are conducted, reducing statistical power due to the p-value adjustment used to limit type I error.

::: callout-note
## `pairs()`

This function conducts all pairwise tests for the specified variable specified. The default p-value adjustment in the `pairs()` function is "tukey"; other options include “holm”, “hochberg”, “BH”, “BY”, and “none”. If you are conducting this on a variable with many levels (e.g. more than five), this adjustment can be severe, resulting in very few statistically significant results. To avoid this, consider other tests such as 'compare to a control' or more tailored custom contrasts. 
:::

### Custom contrasts 

While ***emmeans*** provides many options

First, run emmean object 'm2' for nitrogen treatments.

```{r}
m2
```

Now, let's create a vector for each nitrogen treatment in the same order as presented in output from m2. 
```{r}
A1 = c(1, 0, 0, 0)
A2 = c(0, 1, 0, 0)
A3 = c(0, 0, 1, 0)
A4 = c(0, 0, 0, 1)
```

These vectors (A1, A2, A3, A4) represent each nitrogen treatment in the order as presented in the m2 emmeans object. A1, A2, A3, and A4 vectors represent 0.0 cwt, 0.2 cwt, 0.4 cwt, and 0.6 cwt nitrogen treatments, respectively. 

The next step is to create custom contrasts for comparing '0.0cwt' (A1) treatment to  '0.2cwt' (A2), '0.4cwt' (A3), and '0.6cwt' (A4) treatments. 

```{r}
contrast(m2, method = list(A1 - A2) )
contrast(m2, method = list(A1 - A3) )
contrast(m2, method = list(A1 - A4) )
```

The output shows the difference in mean yield between the control and the three N treatments. The results show that yield increased with nitrogen treatments compared to the control (0.0 cwt) irrespective of the oat variety.

In addition to these custom contrast options, we can also use the built-in functions in the **emmeans** package to compare treatments as per our analysis objective. For example, comparing treatments to control or to compare one treatment level to all other groups.

We will start with examaining m1, an emmeans object.
```{r}
m1
```

Let's suppose we want to compare 'Marvellous' with the other varieties in the analysis. We can conduct "treatment v control" contrasts, setting the variety 'Marvellous' as the control with the `ref` argument.

```{r}
contrast(m1, "trt.vs.ctrl", ref = "Marvellous")
```
Alternatively, we can reference the control group by its appearance in the variety order. In m1, 'Golden rain', 'Marvellous', and 'Victory' are ordered as 1, 2, and 3 respectively.^[This follow an alphabetic ordering that can be altered with the `factor()` function.] 

If we want to contrast 'Golden rain' with other varieties, we can do this by either using `trt.vs.ctrl1`  code, or `trt.vs.ctrlk` and referring to group 1. Both code options will generate the same results.

```{r}
contrast(m1, "trt.vs.ctrl1")
contrast(m1, "trt.vs.ctrlk", ref = 1)
```
We can further customize the contrasts by excluding specific treatment groups from the comparison.

we can demonstrate this by printing the m2 object first to check the treatment order: 
```{r}
m2
```
Next, we can conduct a pairwise comparison by excluding 0.2 cwt. 

```{r}
pairs(m2, exclude = 2)
```

Or we can also evaluate consecutive pairwise comparison as follows:
```{r}
contrast(m2, "consec")
```

## Compact letter displays

Compact letter display (CLD) is a widely used tool by researchers to display multiple comparisons among treatment means. However, the implementation of CLDs can be problematic depending on the analysis aim and experiment design. This is discussed in detail in the [Analysis Tips Chapter](/chapters/analysis-tips.qmd#sec-cld_warning). In particular, using CLD to contrast a large number of levels (e.g. 10 or more) often leads to misinterpretation of the results where people assume items not significantly different from each other are equivalent. It is a good practice in any statistical analysis to carefully consider the goals of a study before implementing CLDs. 

The R package ***multcompView*** [-@multcompView] can be used to generate a display where any two means associated with same symbol are not statistically different using the function `cld()`.

Below are examples of CLD for objects m1 (examining variety), m2 (examining nitrogen), and m4 examining variety within each nitrogen group. In the output below, groups sharing a letter in the $.group$ are not statistically different from each other. 

```{r}
cld(m1, alpha = 0.05, Letters = letters)
```

```{r}
cld(m2, alpha = 0.05, Letters = letters)
```
Let's have a look at the CLDs for the interaction effect:

```{r}
(cld3 <- cld(m3, alpha = 0.05, Letters = letters))
```


## Alternative Approaches

1. Equivalence testing

If your experimental goals include evaluating if two or more treatments are the same, or at the very least equivalent, you can use equivalence testing. This requires that researchers first establish a maximum pairwise difference between items that is not considered meaningful for the data set. That is, what is a numerical difference that is not substantial, and if two items differed by that amount or less, the researcher would not consider that to be a difference that truly mattered. This is a value judgement informed by domain expertise. 

For the next example, let's assume that if the mean difference of two groups differ by less than 30, they can be considered equivalent. We will conduct an equivalence test for all pairwise comparisons of nitrogen treatment emmeans (object m2), specifying the difference in the `delta` argument of the function. 

```{r}
cld(m2, delta = 30, adjust = "none")
```

Here, two treatment groups 0.0 cwt and 0.2 cwt, 0.4 cwt and 0.6 cwt can be considered equivalent.

2. Significance sets 



```{r}
cld(m2, signif = TRUE)
```

It's important to note that we cannot conclude that treatment levels with the same letter are equivalent. We can only conclude that they are not different. 

There is a separate branch of statistics, [equivalence testing](https://en.wikipedia.org/wiki/Equivalence_test){.external target="_blank"}, that is for ascertaining if things are sufficiently similar to conclude they are equivalent. 

See @sec-cld_warning for additional warnings about problems with using compact l

## Export emmeans to an external file

The outputs from `emmeans()` or `cld()` objects can be exported by first converting them to a data frame and using any of the usual export functions (e.g. `write.csv()`)

```{r, eval=FALSE}
result_n <- as.data.frame(m1)

write.csv(result_n, "results.csv", row.names = FALSE)
```

## Graphical display of emmeans 

On easy option is to use default plotting functions that accompany the **emmeans** package. 

```{r}
plot(m1)
plot(m3)
pwpp(m1)
```

Eventually, you will want to create customized plots of model estimates. To do this, the means will have to be manually extracted and converted to a dataframe for usage by plotting libraries. The outputs from the cld4 object (yield of variety within each nitrogen level) can be visualized in ggplot, with variety on the x-axis and estimated means of yield on the y-axis. Different N treatments are presented using different colors.

```{r, warning = FALSE}
ggplot(cld4) +
  aes(x = V, y = emmean, color = N) +
  geom_point(position = position_dodge(width = 0.9)) +
  geom_errorbar(mapping = aes(ymin = lower.CL, ymax = upper.CL), 
                              position = position_dodge(width = 1),
                width = 0.1) +
  geom_text(mapping = aes(label = .group, y = upper.CL * 1.05), 
            position = position_dodge(width = 0.8), 
            show.legend = F)+
  theme_bw()+
  theme(axis.text= element_text(color = "black",
                                size =12))
```
Recall: groups that do not differ significantly from each other share the same letter. 

We can also use the `emmip()` function from the emmeans package to examine the trend in the interaction between variety and nitrogen factors.
```{r}
emmip(model1, N ~ V)
```

::: {.callout-note}
## More details on emmeans

The package ***emmeans*** provides extensive functionality for estimating means, confidence intervals, making contrasts and more. To learn more about this functionality, see the [vignettes](https://rvlenth.github.io/emmeans/){.external target="_blank"} for this package. 
:::

## Conclusion
Be cautious with the terms “significant” and “non-significant”, and don’t interpret a “non-significant” result as there is no effect. Follow good statistical practices such as getting the model right first, and using adjusted P-values for appropriately chosen families of comparisons or contrasts.

::: {.callout-note}
## P values, “significance”, and recommendations

P values are often misinterpreted, and the term “statistical significance” can be misleading. Please refer to this [link](https://www.tandfonline.com/doi/full/10.1080/00031305.2016.1154108#d1e167){.external target="_blank"} to read more about basic principles outlined by the American Statistical Association when considering p-values.
:::
