---
title: Marginal Means & Contrasts
execute: 
  eval: false
---

```{r}
library(nlme); library(emmeans); library(performance)
library(lme4)
```



## Estimated Marginal Means

Estimated marginal means are defined as marginal means of a variable across all levels of other variables in a model, essentially giving a "population-level" average.
model predictions over the grid comprising all factor combinations – called the reference grid

In this chapter, we will demonstrate the extended use of the emmeans package to calculate estimated marginal means and contrasts.



```{r, eval=FALSE}
One-way estimated marginal means and plot

library(multcomp)
library(emmeans)

marginal = emmeans(model,
                   ~ Location)

CLD = cld(marginal,
          alpha=0.05,
          Letters=letters,
          adjust="tukey")

CLD


Location        emmean        SE df  lower.CL upper.CL .group
 Olympia      8.333333 0.6718548 16  6.449596 10.21707  a   
 Northampton 11.833333 0.6718548 16  9.949596 13.71707   b  
 Ventura     13.333333 0.6718548 16 11.449596 15.21707   b  
 Burlington  21.833333 0.6718548 16 19.949596 23.71707    c 



### Order the levels for printing

CLD$Location = factor(CLD$Location,
                       levels=c("Olympia", "Ventura", "Northampton", "Burlington"))

###  Remove spaces in .group 

CLD$.group=gsub(" ", "", CLD$.group)


### Plot

library(ggplot2)

ggplot(CLD,
       aes(x     = Location,
           y     = emmean,
           label = .group)) +

    geom_point(shape  = 15,
               size   = 4) +

    geom_errorbar(aes(ymin  =  lower.CL,
                      ymax  =  upper.CL),
                      width =  0.2,
                      size  =  0.7) +

    theme_bw() +
    theme(axis.title   = element_text(face = "bold"),
          axis.text    = element_text(face = "bold"),
          plot.caption = element_text(hjust = 0)) +

    ylab("Estimated marginal mean\nmidichlorian count") +
    ggtitle ("Midichlorian counts",

             subtitle = "In four U.S. cities") +

            labs(caption  = paste0("\nMidichlorian counts for four locations. ",
                                   "Boxes indicate the EM mean. \n",
                                   "Error bars indicate the 95% ",
                                   "confidence interval of the EM mean. \n",
                                   "Means sharing a letter are not ",
                                   "significantly different (Tukey-adjusted \n",
                                   "comparisons)."),
                            hjust=0.5) +

  geom_text(nudge_x = c(0, 0, 0, 0),
            nudge_y = c(4, 4, 4, 4),
            color   = "black")
```


### Interactions using Emmeans

```{r}

```

```{r, eval=FALSE}
Interaction plot of estimated marginal means

library(multcomp)
library(emmeans)

marginal = emmeans(model,
                   ~ Tribe:Location)

CLD = cld(marginal,
          alpha=0.05,
          Letters=letters,
          adjust="tukey")

CLD


Tribe  Location       emmean        SE df  lower.CL upper.CL .group
 Sith  Olympia      4.333333 0.9501462 16  1.354477  7.31219  a   
 Jedi  Northampton  8.666667 0.9501462 16  5.687810 11.64552  ab  
 Sith  Ventura     10.666667 0.9501462 16  7.687810 13.64552   bc 
 Jedi  Olympia     12.333333 0.9501462 16  9.354477 15.31219   bcd
 Sith  Northampton 15.000000 0.9501462 16 12.021143 17.97886    cd
 Jedi  Ventura     16.000000 0.9501462 16 13.021143 18.97886     d
 Jedi  Burlington  20.666667 0.9501462 16 17.687810 23.64552      e
 Sith  Burlington  23.000000 0.9501462 16 20.021143 25.97886      e



### Order the levels for printing

CLD$Location = factor(CLD$Location,
                       levels=c("Olympia", "Ventura", "Northampton", "Burlington"))

CLD$Tribe = factor(CLD$Tribe,
                       levels=c("Jedi", "Sith"))

###  Remove spaces in .group 

CLD$.group=gsub(" ", "", CLD$.group)


CLD


### Plot

library(ggplot2)

pd = position_dodge(0.4)    ### How much to jitter the points on the plot

ggplot(CLD,
       aes(x     = Location,
           y     = emmean,
           color = Tribe,
           label = .group)) +

    geom_point(shape  = 15,
               size   = 4,
             position = pd) +

    geom_errorbar(aes(ymin  =  lower.CL,
                      ymax  =  upper.CL),
                      width =  0.2,
                      size  =  0.7,
                      position = pd) +

    theme_bw() +
    theme(axis.title   = element_text(face = "bold"),
          axis.text    = element_text(face = "bold"),
          plot.caption = element_text(hjust = 0)) +

    ylab("Estimated marginal mean\nmidichlorian count") +
     ggtitle ("Midichlorian counts for Jedi and Sith",
            subtitle = "In four U.S. cities") +
 
            labs(caption  = paste0("\nMidichlorian counts for two tribes across ",
                                   "four locations. Boxes indicate \n",
                                   "the EM mean. ",
                                   "Error bars indicate the 95% confidence ",
                                   "interval ",
                                    "of the EM \n",
                                   "mean. Means sharing a letter are ",
                                   "not significantly different \n",
                                   "(Tukey-adjusted comparisons)."),
                            hjust=0.5) +
 
  geom_text(nudge_x = c(0.1, -0.1, 0.1, -0.1, 0.1, -0.1, -0.1, 0.1),
            nudge_y = c(4.5,  4.5, 4.5,  4.5, 4.5 , 4.5,  4.5, 4.5),
            color   = "black") +
 
  scale_color_manual(values = c("blue", "red"))


```

### Contrasts using Emmeans
```{r, eval=FALSE}
(warp.emm <- emmeans(warp.lm, ~ tension | wool))

contrast(warp.emm, "poly")
```



```{r}

```





Compact letter display
```{r}
#library(multcomp); library(multcompView)
```

```{r}
#cld(m1, Letters= letters)
```


The letters indicating significant differences can be generated using cld() function from the 'multcomp' package". In the output below, groups sharing a letter in the .group are not statistically different from each other.

::: {.callout-tip}
## Cautionary Note about CLD

It's important to note that we cannot conclude that treatment levels with the same letter are the same. We can only conclude that they are not different. 

There is a separate branch of statistics, "[equivalence testing](https://en.wikipedia.org/wiki/Equivalence_test)" that is for ascertaining if things are sufficiently similar to conclude they are equivalent. 

See @sec-cld_warning for additional warnings about problems with using compact letter display. 
:::



