# Repeated measures mixed models

```{r, include=FALSE, echo=FALSE}
source(here::here("settings.r"))
```

*Check if lmerTest works with lme4 v2.0*

## Background

In the previous chapters, we have covered how to run linear mixed models for various experimental designs. All the examples in those chapters were independent measure designs, where each subject was assigned to a different treatment. Now we will move on to experiment with repeated measures effects (also called "longitudinal data").

Studies that involve repeated data collection from the same experimental units (or subjects) requires a repeated measures component in analysis to model correlations across time for each subject. This is common in studies that are evaluated across different time periods. For example, if samples are collected over different time periods from the same subject or experimental unit, we must model the repeated measures covariance structure when analyzing the treatment effects.

In these models, the 'iid' assumption (independently and identically distributed) is often violated, so we need to introduce specialized covariance structures that can account for these correlations between error terms.

For this these models, we will use ***nlme*** and ***lme4***. These packages provide well documented implementations of heterogeneous error structures. Another package, [***mmrm***](https://CRAN.R-project.org/package=mmrm){.external target="_blank"}, does provide functionality for repeated measures for limited experimental designs, but we will not be exploring that due to limitations in implemented models.  Until recently, implementing hetergenous error terms was [quite challenging in lme4](https://bbolker.github.io/mixedmodels-misc/notes/varmats.html){.external target="_blank"}. Starting with version 4.0 of lme4, functionality for implementing some [covariance structures](https://cran.r-project.org/web/packages/lme4/vignettes/covariance_structures.html) is available. 

::: callout-information

The package [***glmmTMB***](https://CRAN.R-project.org/package=glmmTMB){.external target="_blank"} (generalized linear mixed model using the template model builder) will also implement [many covariance structures](https://cran.r-project.org/web/packages/glmmTMB/vignettes/covstruct.html){.external target="_blank"} and despite its name, can work fine with general linear models (Gaussian) and does not require a mixed component. The implemented structures include compound symmetry, AR1, toeplitz, unstructured and several spatial covariance structures. Truly, this package is incredible. 

:::


In this chapter, we will analyze data from different experiment designs with repeated measures including randomized complete block design, split-plot, and split-split plot designs.

There are several types of covariance structures:

| Structure name | nlme function  | lme4 function | assumption |
|------------------------|------------------------|------------------------|
| Autoregressive (AR1) | `corAR1()` | `ar1()` | observations which are more proximate are more correlated than measures that are more distant |
| Compound symmetry | `corCompSymm()` |  `cs()` | Correlation is equal for all time gaps between observations |
| Unstructured | `CorSymm()` |  `us()`  | Correlations are different for the various time gaps |

::: column-margin
To read more about selecting appropriate covariance structure based on your data, please refer to this [SAS guide](https://support.sas.com/resources/papers/proceedings/proceedings/sugi30/198-30.pdf){.external target="_blank"}.
:::

The repeated measures syntax in **nlme** follow this convention: 

`corAR1(value = 0, form = ~ t|g, fixed = FALSE)`. 

The argument for `value` is the starting value for iterations (zero is the default), and if `fixed = FALSE` (the current nlme default), this value will be allowed to change during the model fitting process. The argument structure for `form`, `~ t|g`, where $t$ is the temporal variable and $g$ is the grouping factor (the subject or experimental unit) to estimate covariance across. Technically,  $g$ is optional and if not specified, the order of the data set will be assumed $t$. However, for most replicated experiments, you will need to specify $g$.  When we use `~t|g` form, the correlation structure is supposed to apply only to observations within the same grouping level. The covariate for this correlation structure must be an **integer value**. Similarly, `corCompSymm()` and `corSymm()` follow the same argument syntax. 

There are other covariance structures enabled by ***nlme*** (e.g., `corARMA()`, `corCAR1()`), but we have found that `corAR1()` and `corCompSymm()` work for most circumstances. Read the [nlme manual](https://cran.r-project.org/web/packages/nlme/refman/nlme.html){.external target="_blank"} if you want to learn more about covariance structures in that package. 

# Examples 

let's start with loading the required libraries for this analysis.

::: panel-tabset
### nlme
```{r, message=FALSE, warning=FALSE}
library(nlme); library(performance); library(emmeans)
library(dplyr); library(broom.mixed); library(ggplot2)
```

### lme4
```{r, message=FALSE, warning=FALSE}
library(lme4); library(performance); library(emmeans)
library(dplyr); library(broom.mixed); library(ggplot2)
```
:::


## RCBD Repeated Measures

First, we will start with the first example with randomized complete block design with repeated measures. 

The data for this example are a sorghum trial, arranged in a RCBD with five replications and 4 levels of the treatment, 'variety'. The dependent variable, y (leaf are index) was measured weekly for 5 weeks. 

We need to have time as both a numeric and a factor variable. In the model, to assess the week effect, week was used as a factor (factweek). For the correlation matrix, the variable "factweek" needs to be converted to a numeric format ("week").

```{r, eval=FALSE, echo=FALSE}
#save data from the package
repeated_meas <- agriTutorial::sorghum %>%   
  mutate(week = as.numeric(factweek),
         block = as.character(varblock)) 

write.csv(repeated_meas, here::here("data", "sorghum.csv"), row.names = FALSE)
```

```{r}
sorghum <- read.csv(here::here("data", "sorghum.csv")) |> 
  mutate(block = as.character(varblock),
         factweek = as.character(factweek),
         variety = as.character(variety))
```

|           |                                     |
|-----------|-------------------------------------|
| block     | blocking unit                       |
| Replicate | replication unit                    |
| Week      | Time points when data was collected |
| variety   | treatment factor, 4 levels          |
| y         | leaf area index                        |

: Table of variables in the data set {tbl-rcbd-rp}

### Data Integrity Checks

Let's do preliminary data check including evaluating data structure, distribution of treatments, number of missing values, and distribution of response variable.

- Check structure of the data

```{r}
str(sorghum)
```

In this data set, 'block', 'factplot', 'factweek' are factors, while 'y' and 'week' are coded as numeric.

- Inspect the independent variables

```{r}
table(sorghum$variety, sorghum$block)
```

The cross tabulation shows a equal number of variety treatments in each block.

- Check the extent of missing data

```{r}
colSums(is.na(sorghum))
```
No missing values 

- Inspect the dependent variable

```{r}
ggplot(data = sorghum, aes(y = y, x = factweek, fill = variety)) +
  geom_boxplot() +  
    theme_bw()
```
It loooks like variety '1' has the lowest yield and showed drastic reduction in yield over weeks compared to other varieties.

One last step before we fit model is to look at the distribution of response variable.

```{r, eval=FALSE}
hist(sorghum$y, main = NA, xlab = "leaf area index")
```

```{r, echo=FALSE}
#| label: fig-rcbd_hist
#| fig-cap: "Histogram of the dependent variable."
#| column: margin
par(mar=c(5.1, 5, 2.1, 2.1))
hist(sorghum$y, main = NA, xlab = "leaf area index", cex.lab = cex_lab, cex.axis = cex_axis, col = base_plot_color)
```

### Model Building

Let's fit the model first using `lme()` from the **nlme** package.

::: panel-tabset
### nlme

```{r}
sorg_1a <- lme(y ~ variety + factweek + variety:factweek,
           random = ~1|block/factplot,
           data = sorghum,
           na.action = na.exclude)
```

### lme4

```{r}
sorg_2a <- lmer(y ~ variety + factweek + variety:factweek + (1|block/factplot),
           data = sorghum,
           na.action = na.exclude)
```
:::

The model fitted above doesn't account for the repeated measures effect. To account for the variation caused by repeated measurements, we can model the correlation among responses for a given subject, which is a plot (factor variable) in this case.

By adding this correlation structure, we are accounting for variation caused by repeated measurements over weeks for each plot. The AR1 structure assumes that data points collected more proximately are more correlated. Whereas, the compound symmetry structure assumes that correlation is equal for all time gaps. Here, we will fit a model with both correlation structures and compare models to find out the best fit model.

In this analysis, the time variable is `week`, and it must be numeric.

In the code chunks above, we fitted two correlation structures including AR1 and compound symmetry matrices. Next we will update the model lm1, with these two matrices. 

::: panel-tabset
### nlme
```{r}
cs1 <- corAR1(form = ~ week|block/factplot,  value = 0.2, fixed = FALSE)
cs1 <- corCompSymm(form = ~ week|block/factplot,  value = 0.2, fixed = FALSE)

sorg_1b <- update(sorg_1a, corr = cs1)
sorg_1c <- update(sorg_1a, corr= cs2)
```

### lme4

```{r}
sorg_2b <- lmer(y ~ variety + factweek + variety:factweek + ar1(0 + week|block/factplot),
           data = sorghum, na.action = na.exclude)
sorg_2c <- lmer(y ~ variety + factweek + variety:factweek + cs(1 + week|block/factplot),
           data = sorghum, na.action = na.exclude)
```
:::




Now let's compare how model fitness differs among models with no correlation structure (lm1), with AR1 correlation structure (lm2), and with compound symmetry structure (lm3). We will compare these models by using `anova()` or by the `compare_performance()` function from the **performance** library.

::: panel-tabset
### nlme

```{r}
anova(sorg_1a, sorg_1b, sorg_1c)
```

Let's compare the models performance to select a model that fits better.

```{r}
#| eval: false
#| include: false
compare_performance(sorg_1a, sorg_1b, sorg_1c)
```

### lme4

```{r}
anova(sorg_2a, sorg_2b, sorg_2c)
```

:::


We prefer to chose model with lower AIC and BIC values. In this scenario, we will move forward with lm2 model containing AR1 structure.

Let's run a `tidy()` on lm2 model to look at the estimates for random and fixed effects.

::: panel-tabset

### nlme

```{r}
tidy(sorg_1b)
```

### lme4

```{r}
tidy(sorg_2c)
```
:::



### Check Model Assumptions

::: panel-tabset
### nlme

```{r, fig.height=3}
#| message: false
#| warning: false
check_model(sorg_1b,  check = c('linearity','qq', 'reqq'), detrend=FALSE, alpha=0)
```

### lme4

```{r, fig.height=3}
#| message: false
#| warning: false
check_model(sorg_2c,  check = c('linearity','qq', 'reqq'), detrend=FALSE, alpha=0)
```

:::


### Inference

::: panel-tabset
### nlme

```{r}
anova(sorg_1b, type = "marginal")
```

### lme4

```{r}
anova(sorg_2c)
```
:::


The ANOVA table suggests a significant effect of the variety, week, and variety x week interaction effect.

We can estimate the marginal means for variety and week effect and their interaction using `emmeans()` function.

::: panel-tabset
### nlme

```{r}
(mean_1 <- emmeans(sorg_1b, ~ variety))

(mean_2 <- emmeans(sorg_1b, ~ variety*factweek))
```

### lme4
```{r}
(mean_1 <- emmeans(sorg_2c, ~ variety))

(mean_2 <- emmeans(sorg_2c, ~ variety*factweek))
```

:::



::: callout-tip
## Representing time variables

It's common to have two versions of temporary variables in a data set: one coded as numeric and the other being coded as a factor or categorical variable. When modelling, the numeric variable is used for fitting correlation matrix, and a factor/character variable may be used in model statement to evaluate the temporal variable's effect on the response if you choose to treat it a categorical variable. It is also a valid choice to view a temporal variables as numeric assuming it represents a real time interval. Days and weeks are real time intervals, while "early", "middle" and "late" are categorical representations of time. 
:::

## Split Plot Repeated Measures

Recall that we evaluated the split plot design in [**Chapter 7**](split-plot-design.qmd). In this example, we will use the same methodology as in Chapter 7 and update it with a repeated measures component.

Next, let's load an alfalfa intercropping data set. This data is from an irrigation and intercropping experiment which was conducted in southern Idaho. Irrigation is the main plot, intercropping is the split plot, and the in-season alfalfa cutting ("cutting") is the unit for repeated measures. 

```{r}
alfalfa <- read.csv(here::here("data/alfalfa_intercropping.csv"))
```

This example contains yield data in a split-plot design. The yield data was collected repeatedly from the same Reps over 5 Sample_times. In this data set, we have:

|             |                                 |
|-------------|---------------------------------|
| cutting       | time points for data collection   |
| irrigation    | Main plot, 2 levels             |
| plot  |   experimental unit          |
| block       |    replication unit     |
| intercrop | Split plot, 3 levels |
| yield | crop yield |
| row |  spatial position by row |
| col | spatial position by column |

: Table of variables in the data set {tbl-split-plot-rp}

### Data Integrity Checks

- Check structure of the data

First, we need to look at the class of variables in the data set.

```{r}
str(alfalfa)
```

We will now convert the fertilizer and rep to factors. In addition, we need to create a new factor variable (sample_time1) to analyze the time effect.

```{r}
alfalfa <- alfalfa |> 
  mutate(cut_num = as.numeric(as.factor(cutting))) |> 
  mutate_at(c("cutting", "irrigation", "plot", "block"), as.factor)
```


To fit the model, we first need to convert Variety, Fertilizer, and Sample_time to factors. In addition, we need to create a variable for each subject which is a 'plot' in this case and contains a unique value for each 'plot'. The plot variable is needed to model the variation in each plot over the sampling time. The plot will be used as a subject with repeated measures. The subject variable can be either a factor or a numeric but the time (could be year, or sample_time) must be a factor.

- Inspect the independent variables


```{r}
table(alfalfa$intercrop, alfalfa$irrigation) 
```
Looks like a balanced design with 2 irrigation treatments and 10 intercropping treatments.

- Check the extent of missing data

```{r}
colSums(is.na(alfalfa))
```

- Inspect the dependent variable

Before fitting a model, let's check the distribution of the response variable.
```{r, echo=FALSE}
#| label: fig-split-plot_hist
#| fig-cap: "Histogram of the dependent variable"
#| column: margin
par(mar=c(5.1, 5, 2.1, 2.1))
hist(alfalfa$yield, main = NA, xlab = "yield", cex.lab = cex_lab, cex.axis = cex_axis, col = base_plot_color)
```

```{r, eval=FALSE}
hist(Yield$Yield, xlab = "yield", main = NA)
```
### Model fit

For this data set, we will use `lme()` from ***nlme*** package and `lmer()` from ***lme4*** to fit a linear mixed model with a repeated measures covariance structure. Instead of the `summary()` function, we will use `tidy()` function from the ***broom.mixed*** package to look at estimates of mixed and random effects. 


::: panel-tabset
### nlme

```{r}
#| warning: false

corr_str1 = corAR1(form = ~ cut_num|block/irrigation/intercrop/plot, value = 0.2, fixed = FALSE)

alf_1 <- lme(yield ~ irrigation*intercrop*cutting,
                random = ~ 1|block/irrigation/intercrop/plot,
                corr= corr_str1,
                data = alfalfa, na.action= na.exclude)
```

### lme4

```{r}
alf_2 <- lme(yield ~ irrigation*intercrop*cutting + 
               ar1(0 + 1|block/irrigation/intercrop/plot),
                data = alfalfa, na.action= na.exclude)
```
:::


### Check Model Assumptions

We will use `check_model()` from the **performance** package to evaluate the model fitness of model fitted using nlme (mod1). 

::: panel-tabset
### nlme

```{r, fig.height=3}
#| warning: false
#| message: false
check_model(alf_1, check = c('linearity', 'qq'), detrend=FALSE, alpha = 0)
```

### lme4

```{r, fig.height=3}
#| warning: false
#| message: false
check_model(alf_2, check = c('linearity', 'qq'), detrend=FALSE, alpha = 0)
```

:::

These diagnostic plots look fine. The linearity and homogeneity of variance plots show no trend. The Q-Q plots for the overall residuals and for the random effects fall on a straight line so we can be satisfied with that.

### Inference

::: panel-tabset
### nlme

```{r}
#| warning: false
#| message: false
anova(alf_1, type = "marginal")
```

### lme4

```{r}
#| warning: false
#| message: false
anova(alf_2)
```
:::

 
Next, we can estimate marginal means and confidence intervals at different levels of the independent variables using `emmeans()`.

```{r}
emmeans(alf_1,~ cutting)
emmeans(alf_1,~ intercrop)
```

### lme4

```{r}
emmeans(alf_2,~ cutting)
emmeans(alf_2,~ intercrop)
```
:::

::: column-margin
To explore more about contrasts and emmeans please refer to [**Chapter 13**](means-and-contrasts.qmd).
:::

## Split-split Plot Repeated Measures {#ssp-rm}

Recall, we have evaluated this split-split plot experiment design in the [**split-split design chapter**](split-split-plot-design.qmd). In this example,a repeated measures component will be added to the split-split plot model.

```{r}
phos <- read.csv(here::here("data", "split_split_repeated.csv"))
```

|             |                                 |
|-------------|---------------------------------|
| plot        | experimental unit                |
| block     | replication unit            |
| Ptrt  | Main plot, 2 levels           |
| Inoc       | Split plot, 2 levels         |
| Cv    | Split-split plot, 5 levels
| time | time points for data collection |
| P_leaf | leaf phosphorous content  |

### Data Integrity Checks

- Check structure of the data

```{r}
str(phos)
```

We need two variables for time, one formatted as a factor and the other numeric.

```{r, warning=FALSE, message=FALSE}
phos1 <- phos %>%   
  mutate(
    time = as.factor(time), 
    time_num = as.numeric(time),
    rep = as.character(bloc),
    plot = as.character(plot)) 
```

- Inspect the independent variables

```{r}
table(phos1$Cv, phos1$Ptrt, phos1$Inoc) 
```
Looks like a well balanced design with 2 variety treatments and 3 fertilizer treatments.

- Check the extent of missing data

```{r}
colSums(is.na(phos1))
```
No missing values in data. 

- Inspect the dependent variable

Before fitting a model, let's check the distribution of the response variable.
```{r, echo=FALSE}
#| label: fig-split-split-plot_hist
#| fig-cap: "Histogram of the dependent variable."
#| column: margin
par(mar=c(5.1, 5, 2.1, 2.1))
hist(phos1$P_leaf, main = NA, xlab = "P leaf", cex.lab = cex_lab, cex.axis = cex_axis, col = base_plot_color)
```

```{r, echo=FALSE}
#| label: fig-split-split-plot_boxplot
#| fig-cap: "Boxplot of P concentration by stage"
#| column: margin
par(mar=c(5.1, 5, 2.1, 2.1))
```

```{r, eval=FALSE}
hist(phos1$P_leaf, main = NA, xlab = "P leaf")
boxplot(P_leaf ~ time, data = phos1, main = NA)
```

::: callout-warning
## distribution of dependent variables

Note that we observed an uneven distribution of response variable with a bimodal distribution and a noticeable gap in the 500 to 1500 range. Given this odd distribution, it may be tempting to consider a transformation in order to attempt to impose normality. It's important to remember that the assumption of normality applies to the residuals, not the raw data. Plotting the data is for checking the data looks *as expected*, a judgment that requires some knowledge of the experiment (this was Julia Piaskowski's PhD research). In this case, the time points, PT1 and PT2, correspond to early wheat growth stages (tillering and jointing, respectively), and the final time point represents senescent leaf tissue at grain maturity. At that physiological stage, it is normal for phosphorus leaf concentration to be much lower. Since the data look as expected, we will proceed with a general linear model and evaluate the residuals from the model-fitting process when deciding if a non-normal distribution is appropriate for the data.  
:::

### Model fit

::: panel-tabset
### nlme

```{r}
#| warning: false
corr_str1 = corCompSymm(form = ~ time_num|rep/Ptrt/Inoc/plot, value = 0.2, fixed = FALSE)

pfit1 <- lme(P_leaf ~ time*Ptrt*Inoc*Cv,
                random = ~ 1|rep/Ptrt/Inoc/plot,
                corr = corr_str1,
                data = phos1, na.action= na.exclude)

```

### lme4

```{r}
pfit2 <- lmer(P_leaf ~ time*Ptrt*Inoc*Cv + cd(1|rep/Ptrt/Inoc/plot),
                data = phos1, na.action= na.exclude)
```
:::

### Check model assumptions 

::: panel-tabset
### nlme

```{r, fig.height=3}
#| warning: false
#| message: false
check_model(pfit1, check = c('linearity', 'qq'), detrend=FALSE, alpha =0)
```

### lme4

```{r, fig.height=3}
#| warning: false
#| message: false
check_model(pfit2, check = c('linearity', 'qq'), detrend=FALSE, alpha =0)
```

:::

This model fit a first glance is not ideal, but that LOESS line is trying to model a space where there are no data (between 500 and 1500 ppm P leaf concentration), so that can introduce artifacts. **Performance** does have an option for testing for heteroscedascity:

::: panel-tabset
### nlme

```{r}
check_heteroscedasticity(pfit1)
```

These results do confirm our suspicions that the residuals were not as heteroscedastic as they first appeared. However, the boxplot indicated a difference in variance for each time point. This issue is addressed in the [chapter on variance components](variance-components.qmd). 

```{r}
#| include: false
#| echo: false
#| eval: false
save(pfit1, corr_str1, file = here::here("data/phos_fit1.RData"))
```

### lme4

```{r}
check_heteroscedasticity(pfit2)
```
:::


### Inference

::: panel-tabset
### nlme

```{r}
#| warning: false
#| message: false
anova(pfit1, type = "marginal")
```

### lme4

```{r}
#| warning: false
#| message: false
anova(pfit2)
```
:::

::: panel-tabset
### nlme

```{r}
emmeans(pfit1, ~ Inoc|Cv)
emmeans(pfit1, ~ time|Cv)
```

### lme4

```{r}
emmeans(pfit2, ~ Inoc|Cv)
emmeans(pfit2, ~ time|Cv)
```
:::

