
#Strip Plot Design

```{r, include=FALSE, echo=FALSE}
source(here::here("settings.r"))

par(mar=c(5.1, 6, 4.1, 2.1))
```

## Background

It is very common design where experimental treatments are applied at random to experimental units within each block. The blocks are intended to control for a nuisance source of variation, such as over time, spatial variance, changes in equipment or operators, or myriad other causes.

The statistical model:

::: panel-tabset
### lme4
```{r, message=FALSE, warning=FALSE}
library(lme4); library(lmerTest); library(emmeans)
library(dplyr); library(performance); library(ggplot2)
library(broom.mixed)
```
### nlme

```{r, message=FALSE, warning=FALSE}
library(nlme); library(performance); library(emmeans)
library(dplyr); library(ggplot2); library(broom.mixed)
```
:::


A strip-plot experiment with three reps, variety as the horizontal strip and nitrogen fertilizer as the vertical strip.
```{r}
data1 <- agridat::gomez.stripplot

```

```{r}
str(data1)
```

```{r}
library(desplot)
desplot(data1, gen ~ col*row,
        # aspect unknown
        out1=rep, out2=nitro, num=nitro, cex=1,
        main="Strip Plot Design")
```


```{r}
data1$nitro <- as.factor(data1$nitro)
```


```{r}

```



