---
title: Variance & Variance Components
execute: 
  eval: false
---

Mixed models provide the advantage of being able to estimate the variance of random variables. Instead of looking at a variable as a collection of specific levels to estimate, random effects view variables as being a random drawn from a normal distribution with a standard deviation. The decision of how to designate a variable as random or fixed depends on 

```{r}
library(nlme); library(emmeans); library(performance)
library(lme4)
```


## Unequal Variance

### Case 1: Unequal Variance Due to a Factor

```{r}
var_ex1 <- here::here(read.csv("data", "MET_trial_variance.csv"))
```

```{r}
var_ex1$block <- as.character(var_ex1$block)
hist(var_ex1$yield)
boxplot(yield ~ site, data = var_ex1)
```

```{r}
m1_a <- lme(yield ~ site:variety + variety, 
                random = ~ 1 |site/block, 
                na.action = na.exclude, 
                data = var_ex1)

```


```{r}
m1_b <- update(m1_a, weights = varIdent(form = ~1|site))
```

::: {.note}


```{r eval=FALSE}
m1_b <- update(m1_a, weights = varIdent(form = ~1|site))
```

is equivalent to

```{r eval=FALSE}
m1_b <- lme(yield ~ site:variety + variety, 
                random = ~ 1 |site/block,
                weights = varIdent(form = ~1|site), 
                na.action = na.exclude, 
                data = var_ex1)
```

:::

### Case 2: Variance is related to the fitted values

```{r}
var_ex2 <- read.csv(here::here("data", "single_trial_variance.csv"))
```

```{r}
var_ex1$block <- as.character(var_ex1$block)
hist(var_ex2$yield)
```


```{r}
m2_a <- lme(yield ~ variety, 
               random = ~ 1 |block, 
               na.action = na.exclude, 
               data = var_ex2)
```

```{r}
m2_b <- update(m2_a, weights = varPower())
```

## Coefficient of Variation

```{r}
m2_ave <- fixef(m2_b)[1]
names(m2_b) <- NULL
```

```{r}
m2_cv = sigma(m2_b)/m2_ave*100
m2_cv
```



### Looking at Variance Components

```{r}
var_comps <- read.csv(here::here("data", "potato_tuber_size.csv"))
```




